\propertitle

### 2019-12-02 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-02T14:16:29+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal population\\ from subpopulation data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

The first and most important assumption of this kind of maximum-entropy
application in neuroscience can be expressed in three equivalent ways:
\begin{enumerate*}[(\roman*)]
\item the order of the time bins has no relevance; \item the frequencies of
  the activity during a recording are informationally sufficient for
  inferences about unrecorded times; \item our degree of belief about the
  sequence of activities is invariant under permutations of the time bins,
  that is, exchangeable \citep[\chap~4]{bernardoetal1994_r2000}{dawid2013}.
\end{enumerate*}
Equating time averages and expectations, \eqn~\eqref{eq:final_constraint},
would not make sense without this assumption. Its mathematical consequence
\citep{definetti1930,hewittetal1955} is that the probability of observing a
frequency distribution $(\yFF_{\yA})$ in $T$ time bins is given by a
mixture of terms
\begin{equation}
  \label{eq:mixture_from_definetti}
  \exp[-T\;\sh(\yF;\ynu)] \quad
  \text{with weights }q(\ynu),
\end{equation}
where $\ynu\defd (\nu_{\yA})$ represents the long-run frequency
distribution for the activity. % Other maximum-entropy models are also used
% in neuroscience, such as kinetic Ising models
% \cite{kawasaki1972,mezardetal2011,battistinetal2015}, based on the slightly
% less drastic assumption of invariance under permutations of sequences
% having the same starting state and the same transition counts
% \citep{diaconisetal1980c,geetal2012}.
Note that the maximum-entropy method in statistical physics is meant to be
used not with this assumption, but with averages across \emph{repetitions}
of the same experiment: its constraints represent macroscopically
\emph{reproducible} quantities and kinematics, even in non-equilibrium
\citep{jaynes1957,guntonetal1983,tikochinskyetal1984,grandy1988,berryetal1988,deroecketal2005_r2006,touchette2009}.

The second assumption is specific to our application: all subsets of $\yA$
neurons in the larger population are equally likely to give rise to total
activity $\yA$, at every time bin. A mathematically equivalent assumption
is that $n$ neurons are sampled anew at each time bin.
\begin{equation}
  \label{eq:assume_sampled_anew}
  \pxx[(\ya_{1},\dotsc,\ya_{T}), (\yA_{1}, \dotsc, \yA_{T})]
  = \prod_{t=1}^{T} \pxx(\ya_{t}, \yA_{t})
  = \prod_{t=1}^{T} \yG_{\ya_{t}\yA_{t}}\;\nu_{\yA_{t}}
\end{equation}

Both assumptions above make sense as toy hypotheses for judging the
informational relevance of correlations of some order, as in
\sect~\ref{sec:marginalization}, to be compared with hypotheses about more
realistic time dependences and clustering.

In an exact calculation based on the two assumptions above, the probability
of the sequences of sample and larger-population activities,
$(\ya_{t}, \yA_{t})$, involves the joint frequencies of total activities of
sample and larger population $J_{\ya,\yA}$, multiplied by the number of
time bins $T$:
\begin{equation}
  \label{eq:prob_joint_frequencies}
  \underbracket{\prod_{t} \pxx(\ya_{t}, \yA_{t})}_{\mathclap{\textit{prob. of joint sequence}}} \equiv
  \prod_{\ya,\yA} \pxx(\ya, \yA)^{T\,J_{\ya,\yA}} =
  \prod_{\ya,\yA}\bigl[\underbracket{\vphantom{\underline{g}}\yG_{\ya\yA}\;\nu_{\yA}}_{\mathclap{\textit{sampling relation}}}\bigr]^{T\,J_{\ya,\yA}}.
\end{equation}
The reasoning would be similar to that leading to
\eqn~\eqref{eq:prob_frequencies} of appendix~\ref{sec:measure_suff}.
Conditioning on the frequencies $\yf$ of the sample would correspond to
constraining the marginal $\sum_{\yA}F_{\ya,\yA}=\yff_{\ya}$. The
maximum-entropy method approximates $T$ with infinity, thus treating the
probability above as a delta. The result is that
\begin{enumerate*}[label=(\roman*)]
\item the probability and observed frequency for the total activity $\yA$
  are constrained to be equal,
\item the frequency of sample activity $\ya$ conditional on
  larger-population activity $\yA$ is constrained to equal the
  hypergeometric distribution $\yG_{\ya\yA}$,
  \eqn~\eqref{eq:conditional_hypergeometric}.
\end{enumerate*}
Without this approximation the latter conditional frequency can differ from
$\yG_{\ya\yA}$, and this is expected if the number of bins $T$ is not much
larger than the number of possible joint possibilities, roughly
$n\times N$.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-02T15:04:31+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal population\\ from subpopulation data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

The first and most important assumption of this kind of maximum-entropy
application in neuroscience can be expressed in three equivalent ways:
\begin{enumerate*}[(\roman*)]
\item the order of the time bins has no relevance; \item the frequencies of
  the activity during a recording are informationally sufficient for
  inferences about unrecorded times; \item our degree of belief about the
  sequence of activities is invariant under permutations of the time bins,
  that is, exchangeable \citep[\chap~4]{bernardoetal1994_r2000}{dawid2013}.
\end{enumerate*}
Equating time averages and expectations, \eqn~\eqref{eq:final_constraint},
would not make sense without this assumption. Its mathematical consequence
\citep{definetti1930,hewittetal1955} is that the probability of observing a
sequence of activities $(\yA_{1},\dotsc,\yA_{T})$ with frequencies
$(\yFF_{\yA})$ is a mixture of terms
\begin{equation}
  \label{eq:mixture_from_definetti}
%  \exp[-T\;\sh(\yF;\ynu)] \quad
%  \pxxx[(\yA_{1},\dotsc,\yA_{T})] \propto
  \prod_{\yA=1}^{N}{\nu_{\yA}}^{T\,\yFF_{\yA}} \quad
  \text{with weights }q(\ynu),
\end{equation}
where $\ynu\defd (\nu_{\yA})$ represents the long-run frequency
distribution for the activity. % Other maximum-entropy models are also used
% in neuroscience, such as kinetic Ising models
% \cite{kawasaki1972,mezardetal2011,battistinetal2015}, based on the slightly
% less drastic assumption of invariance under permutations of sequences
% having the same starting state and the same transition counts
% \citep{diaconisetal1980c,geetal2012}.
Note that the maximum-entropy method in statistical physics is meant to be
used not with this assumption, but with averages across \emph{repetitions}
of the same experiment: its constraints represent macroscopically
\emph{reproducible} quantities and kinematics, even in non-equilibrium
\citep{jaynes1957,guntonetal1983,tikochinskyetal1984,grandy1988,berryetal1988,deroecketal2005_r2006,touchette2009}.

The second assumption is specific to our application: all subsets of $\yA$
neurons in the larger population are equally likely to give rise to total
activity $\yA$, at every time bin. A mathematically equivalent assumption
is that $n$ neurons are sampled anew at each time bin. This assumption
allows us to factorize the joint probability of the two sequences of
activities and apply the sampling
formula~\eqref{eq:conditional_hypergeometric} at each bin:
\begin{equation}
  \label{eq:assume_sampled_anew}
  \pxx[(\ya_{1},\dotsc,\ya_{T}) \| (\yA_{1}, \dotsc, \yA_{T}) ]
  = \prod_{t=1}^{T} \pxx(\ya_{t} \| \yA_{t})
  = \prod_{t=1}^{T} \yG_{\ya_{t}\yA_{t}}.
\end{equation}

Both assumptions above make sense as reference hypotheses for judging the
informational relevance of correlations of some order, as in
\sect~\ref{sec:marginalization}, to be compared with hypotheses about more
realistic time dependences and clustering.

From these two assumptions the probability calculus allows us to calculate
the probability for the frequency distribution $\yF$ given the observed
frequency distribution for the sample $\yf$. The result is a mixture of
terms involving the joint frequencies of total activities of sample and
larger population $J_{\ya,\yA}$, multiplied by the number of time bins $T$:
\begin{equation}
  \label{eq:prob_joint_frequencies}
%\pxx[\yF \| \yf ] \propto
\prod_{\ya,\yA}\bigl(\yG_{\ya\yA}\;\nu_{\yA}\bigr)^{T\,J_{\ya,\yA}},
\end{equation}
with weights $q(\ynu)$, and with the marginal of the joint frequency
constrained Conditioning on the frequencies $\yf$ of the sample
would correspond to constraining the marginal
$\sum_{\yA}F_{\ya,\yA}=\yff_{\ya}$. The maximum-entropy method approximates
$T$ with infinity, thus treating the probability above as a delta. The
result is that
\begin{enumerate*}[label=(\roman*)]
\item the probability and observed frequency for the total activity $\yA$
  are constrained to be equal,
\item the frequency of sample activity $\ya$ conditional on
  larger-population activity $\yA$ is constrained to equal the
  hypergeometric distribution $\yG_{\ya\yA}$,
  \eqn~\eqref{eq:conditional_hypergeometric}.
\end{enumerate*}
Without this approximation the latter conditional frequency can differ from
$\yG_{\ya\yA}$, and this is expected if the number of bins $T$ is not much
larger than the number of possible joint possibilities, roughly
$n\times N$.
